---
title: "Younger age groups, women, and lower levels of education are more likely to vote Democractic"
subtitle: "A logistic regression analysis of demographics in the 2020 US Presidential Election"
author: 
  - Victor Ma
thanks: "Code and data are available at: https://github.com/bestmustard/Determinants-of-Political-Support-in-the-United-States."
date: 03/16/2024
date-format: long
abstract: "In this study, I investigate demographic patterns which influence political support for the US 2020 election, specifically focusing on whether demographics of people tend to vote for Joe Biden or Donald Trump. I use a logistic regression model with the variables age, gender, and education to predict which categories in each demographic have the highest chance of voting for Joe Biden. I find that the model accurately predicts the trend of younger people, women, and those with the lowest level of education being more likely to support Joe Biden."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
    
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dataverse)
library(knitr)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(dplyr)
library(patchwork)
library(ggplot2)
library(broom.mixed)
library(GGally)

ces_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

# Introduction

The 2020 United States presidential election was a landmark event that encapsulated a confluence of crises and societal shifts. Beyond the COVID-19 pandemic, which reshaped healthcare priorities and exposed systemic vulnerabilities, the election underscored debates about economic inequality, particularly as millions faced unemployment during the economic downturn [@Blustein2020; @BLS2020]. Climate change emerged as a defining issue, with voters increasingly prioritizing environmental policies in response to intensifying natural disasters [@IPCC2020].

Demographic shifts also played a critical role. The increasing influence of younger, more diverse voters highlighted generational and racial divides in political priorities [@Frey2020]. Meanwhile, changes in suburban voting patterns, particularly among college-educated women, illustrated the evolving dynamics of key swing areas [@Cook2020].

Another major factor was the unprecedented role of mail-in voting, which became a focal point for controversy amidst the pandemic and claims of election fraud. This shift in voting methods not only increased accessibility for millions but also fueled political and legal battles over election integrity [@NCSL2020]. Finally, the election was a litmus test for the resilience of democratic institutions in the face of misinformation, political polarization, and challenges to the legitimacy of results [@BenGurion2020; @Sunstein2020].

Analyzing how living environments, employment status, and income levels correlate with political alignment in 2020 offers deep insights into societal trends. Research shows that urban residents tend to favor liberal candidates, reflecting the diversity and progressive social policies often prioritized in cities [@bishop2019bigsort]. Conversely, rural voters strongly supported Donald Trump, consistent with longstanding trends emphasizing traditional values and conservative policies [@cramer2016politics].

Employment status also played a key role. Data from the Cooperative Congressional Election Study (CCES, 2020) revealed that students and retirees, often focused on issues like education and healthcare, leaned more toward Joe Biden, while full-time workers, prioritizing economic stability, showed mixed preferences based on industry and region [@cces2020].

Income disparities further highlighted political divides. Lower-income groups, grappling with economic inequality, largely supported Biden due to his policies on healthcare and social safety nets, as noted in Pew Research’s 2020 voter analysis. Higher-income voters, on the other hand, leaned Republican, particularly in rural and suburban settings, reflecting preferences for tax policies that align with their financial interests [@pew2020voteranalysis].

These findings display the imapct of personal circumstances and political preferences, mapping not only the political landscape of 2020 but also providing a deeper understanding of the evolving priorities shaping voter behavior.

In this study, I use a logistic regression model to predict which candidate—Joe Biden or Donald Trump—was more likely to receive votes based on factors such as living environment, employment status, and income levels. Logistic regression is particularly suited for this analysis as it models binary outcomes, allowing us to estimate the likelihood of a voter aligning with one candidate over the other. By focusing on these explanatory variables, the study provides valuable insights into how socioeconomic and regional demographics influenced voter preferences during one of the most polarized elections in modern history.

The primary dataset for this study is the 2020 Cooperative Congressional Election Study (CCES), a stratified sample survey administered by YouGov and sourced from the Harvard Dataverse [@cces2020]. 

Complementary insights are drawn from Pew Research Center’s 2020 voter analysis, which contextualizes demographic trends across key groups, and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which provides aggregate statistics on voter turnout and demographic characteristics [@pew2020; @CPS2020]. These additional datasets validate findings from the CCES and enhance the robustness of our conclusions.

My report is structured into five main sections and includes an appendix. After the introduction, the first section describes the CCES 2020 dataset, highlighting key attributes and presenting graphs of demographic distributions. The second section details the logistic regression model, its rationale, and preliminary findings. The third section analyzes how living environment, employment status, and income influenced election outcomes using visualizations and numerical results. The fourth section discusses the implications of the findings, potential weaknesses, and future research directions.

The appendix provides an in-depth exploration of survey methodologies, focusing on sampling and observational data related to the CCES 2020 dataset. It includes detailed discussions of idealized methodologies, literature comparisons, and simulations to illustrate sampling variability and bias, expanding on points in the Data section.

This analysis was conducted in R, utilizing several packages to streamline data handling, visualization, and modeling. Key tools include tidyverse for data manipulation and visualization, dataverse for accessing the CCES dataset, knitr for generating the report, modelsummary for interpreting models, and rstanarm for Bayesian regression analysis [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr]. Additionally, ggplot visualizations and elements of the "Discussion" section were developed with the assistance of ChatGPT4 @chatgpt.
\newpage


# Data
The dataset for this analysis is the 2020 Cooperative Election Study (CCES) [@ces2020], accessed via the Harvard Dataverse. The 2020 iteration surveyed 61,000 respondents, capturing a broad spectrum of U.S. political opinions. The release includes detailed documentation and questionnaires, ensuring transparency and replicability. Vote validation was conducted by Catalist, a large-scale organization maintaining data on over 240 million U.S. individuals [@Catalist2017].

## Strengths

### Sample Size

Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[@Gallup2020], [@ces2020].

```{r out.height="30%"}
#| label: fig-respondents
#| fig-cap: "Number of respondents for various 2020 election survey sources"
#| echo: false
#| warning: false
#| message: false

# Create a data frame with the number of respondents for each source
survey_data <- data.frame(
  Source = c("ANES 2020", "Pew Research", "Edison Exit Polls", "CCES 2020"),
  Respondents = c(8280, 11818, 25000, 61000) # Updated with additional sources
)

# Generate the bar graph
ggplot(survey_data, aes(x = Source, y = Respondents, fill = Source)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Number of Respondents per Survey Source",
    x = "Survey Source",
    y = "Respondents"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Angle the x-axis text for readability

```


@fig-respondents shows that CCES poll features the largest respondent pool compared to other surveys. This extensive sample size enables granular analyses, ensuring the representation of smaller subgroups that may be underrepresented in smaller datasets [@Ansolabehere2020].

### Data Collection Phases
The CCES collected data in two waves, immediately before and after the 2020 election. This helps capture the shifts in voter attitudes and preferences during the critical final stages of the election campaign [@ces2020]. Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture as broad of a perspective as the CES survey [@Pew2020], [@Gallup2020].

### Sampling Methodology
The CCES employs a two-step sample matching methodology designed to create a representative survey sample:

#### Target Sample Creation
A target sample is generated to reflect the broader population based on key demographic variables such as age, gender, race, and educational attainment. This target sample serves as a benchmark for representativeness.

#### Matched Sample Selection
For each individual in the target sample, a closely matched respondent is selected from a pool of survey volunteers maintained by YouGov. Matches are determined using a combination of publicly available databases and YouGov’s proprietary data, ensuring close alignment on variables such as geography, party registration, and voter history.

Once the matched sample is established, it is weighted to adjust for discrepancies between the sample and the target population. The weighting process involves two stages:

#### Demographic Weighting
Adjustments are made using data from the American Community Survey to ensure accurate representation across demographics.

#### Voting Behavior Weighting
Further adjustments incorporate validated voter registration data, accounting for turnout and voting patterns at the state and national levels.

Validation of the sampling methodology is conducted by comparing state-level survey results with official election outcomes. This comparison ensures that survey estimates align with actual voter behavior, providing confidence in the dataset's representativeness.

The sampling approach is specifically designed for online opt-in panels and mitigates biases commonly associated with such surveys, such as overrepresentation of politically engaged respondents. Unlike traditional random or quota sampling, this methodology emphasizes precision in matching and post-stratification to minimize systematic error [@ces2020], [@sample2007].

## Limitations
The data has some limitations. An error affected 925 North Carolina respondents who were shown incorrect House race candidates, impacting analyses of these districts. Additionally, self-reported data can introduce biases such as recall or social desirability bias. Pre-election surveys may also fail to capture last-minute shifts in voter sentiment.

The online nature of the survey, hosted by YouGov, expands accessibility but may underrepresent groups with limited internet access or technological literacy. Furthermore, the lack of an interviewer in online surveys could impact response quality [@ces2020].

## Variables of Interest
The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that underscores their influence on voting behavior.

### Urban vs. Rural Residency
Urban and rural divides are well-documented predictors of political preferences. Urban voters have consistently leaned Democratic, driven by higher population density and exposure to diverse cultural and socioeconomic dynamics. For example, Pew Research found that 62% of urban voters supported Hillary Clinton in the 2016 election, compared to only 35% in rural areas [@Pew2018]. Rural voters, in contrast, exhibit stronger support for Republican candidates, influenced by traditional values and economic concerns rooted in agriculture and resource-based industries [@Cramer2016].

Research also shows that suburban areas, often politically contested, have shifted in recent years. The Cook Political Report highlights a notable swing in suburban support toward Democratic candidates in 2020, attributed to changes in demographics and education levels among suburban voters [@Cook2020].

### Income Level
Income levels are closely tied to political preferences, with higher-income individuals generally favoring Republican candidates due to tax policies, while lower-income groups often lean Democratic, prioritizing social welfare and redistribution policies [@Frank2004]. For example, the Census Bureau's Current Population Survey shows that in 2020, households earning under $50,000 were more likely to support Joe Biden, while those earning over $100,000 favored Donald Trump [@CPS2020].

However, this trend is nuanced by education levels and geographic factors. High-income earners in urban areas often prioritize social liberalism and climate change policies, aligning with Democratic platforms, while rural high-income earners focus more on fiscal conservatism [@Pew2020; @Edsall2020].

### Employment Level
Employment status significantly impacts voter preferences, with distinct patterns emerging across different occupational categories. Full-time workers are often divided along industry lines, with white-collar employees tending to support Democratic candidates and blue-collar workers aligning more with Republican candidates [@Muro2020]. Unemployment during economic crises, such as the COVID-19 pandemic, has further shaped voting patterns. Studies show that unemployed individuals are more likely to support candidates promising expansive social safety nets and job creation [@Blustein2020; @Fowler2020].

Retirees also exhibit unique voting behaviors, often prioritizing stability and healthcare, leading to a higher likelihood of Republican support [@KFF2020]. In contrast, students, who are less economically established and more progressive, tend to favor Democratic candidates [@Frey2020].

## Data Preparation and Cleaning
The 2020 CCES data, as recorded by Schaffner, Brian et al., was obtained from the Harvard Dataverse and processed using the arrow package for efficient storage in parquet format [@arrow]. The raw data was imported from a CSV file, and the cleaning process involved filtering for registered voters who cast votes for either Joe Biden or Donald Trump, treating the presidential vote as a binary outcome.

Key variables were transformed for clarity and analysis:

Living Environment: Categorized into "City," "Suburb," "Town," and "Rural area" based on the urbancity variable.
Employment Status: Labeled as "Full-time," "Part-time," "Temporarily laid off," "Unemployed," "Retired," "Permanently disabled," "Homemaker," or "Student."
Income Levels: Grouped into ranges: "< 10k," "10-50k," "50-100k," "100-200k," "200-500k," and "> 500k."
These transformations were implemented using the dplyr package for data manipulation. Each variable was converted into a factor with ordered levels to facilitate analysis and visualization.

The cleaned dataset was saved in both CSV and parquet formats for ease of use in subsequent analyses. Distributions for key explanatory variables, such as living environment, employment status, and income, are visualized in @fig-living, @fig-employment, and @fig-income below.

```{r out.height="20%"}
#| label: fig-living
#| fig-cap: "Number of respondents by area of living"
#| echo: false
#| warning: false
#| message: false

# Plot for area of living, excluding NA values
ggplot(cleaned_data %>% filter(!is.na(living)), aes(x = living)) +
  geom_bar(fill = "#87CEEB") +
  theme_minimal() +
  labs(title = "Respondents per Area of Living",
       x = "Living Environment",
       y = "Count") +
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 14))

```
```{r out.height="20%"}
#| label: fig-employment
#| fig-cap: "Number of respondents by employment status"
#| echo: false
#| warning: false
#| message: false

# Plot for employment status
ggplot(cleaned_data %>% filter(!is.na(employment)), aes(x = employment)) +
  coord_flip() +
  geom_bar(fill = "#87CEFA") +
  theme_minimal() +
  labs(title = "Respondents per Employment Status",
       x = "Employment Status",
       y = "Count") +
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 14))

```
```{r out.height="20%"}
#| label: fig-income
#| fig-cap: "Number of respondents by family income level"
#| echo: false
#| warning: false
#| message: false

# Plot for income level, excluding NA values
ggplot(cleaned_data %>% filter(!is.na(income)), aes(x = income)) +
  geom_bar(fill = "#FFD700") +
  theme_minimal() +
  labs(title = "Respondents per Income Level",
       x = "Income Level",
       y = "Count") +
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for clarity
```
@fig-living shows that the most common respondents were living in the suburbs and the city. 

In @fig-employment, the most common employment levels were full-time employees or retired.

@fig-income shows that the most common income levels were between the 10-100k range, and there was a high number of participants with an income level above $500,000.

```{r out.height="30%"}
#| label: fig-scatter-income-employment
#| fig-cap: "Income vs. Employment Status"
#| echo: false
#| warning: false
#| message: false

ggplot(cleaned_data, aes(x = employment, y = as.numeric(income), color = voted_for)) +
  geom_jitter(width = 0.2, height = 0) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "Income vs. Employment Status",
    x = "Employment Status",
    y = "Income Level",
    color = "Voted For"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Slant x-axis text
  )

```

@fig-scatter-income-employment shows us the income distribution across employment types.

Full-time workers exhibit the widest income distribution, spanning from "<10k" to ">500k."
Retirees and students predominantly occupy lower income levels, aligning with expectations based on fixed or limited income sources.

Retirees and students show a strong preference for Biden, while higher-income full-time workers lean slightly toward Trump.
Homemakers and permanently disabled individuals display narrower income ranges but include both Biden and Trump voters.

In the ">500k" category, Trump garners more support, consistent with Republican tax policies favoring high-income earners.
These patterns underscore the nuanced relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.

```{r out.height="30%"}
#| label: fig-heatmap-employment-living
#| fig-cap: "Heatmap of employment status and living environment"
#| echo: false
#| warning: false
#| message: false

# Pre-compute counts for the heatmap
heatmap_data <- cleaned_data %>%
  count(living, employment)

# Create the heatmap
ggplot(heatmap_data, aes(x = living, y = employment, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() +
  labs(title = "Employment Status by Living Environment",
       x = "Living Environment",
       y = "Employment Status",
       fill = "Count")
```

The @fig-heatmap-employment-living provides insights into the distribution of employment statuses across different living environments:

Suburbs have the highest concentration of full-time workers, as indicated by the deep blue color. This aligns with the demographic trends of suburban areas, which often attract working professionals due to proximity to urban centers and family-friendly amenities.

Retirees are present in all living environments, with notable concentrations in both suburbs and rural areas. This reflects retirement trends where individuals seek quieter or more affordable living conditions outside cities.

Rural areas show a higher proportion of homemakers compared to other environments, consistent with traditional family roles in less urbanized areas.

Cities have the largest concentration of students, likely due to the presence of educational institutions and urban amenities.

Rural areas show higher proportions of unemployed and permanently disabled individuals compared to cities and suburbs, highlighting economic disparities and limited job opportunities in these regions.

Part-time work and temporary layoffs are distributed fairly evenly across living environments, suggesting they are less influenced by geographic location.

# Model

The goal of this logistic regression model is to predict whether a respondent votes for Joe Biden (1) or Donald Trump (0) based on employment status, income level, and living environment.

The model calculates the log odds of a respondent voting for Joe Biden and maps it to a probability between 0 and 1 using the logistic function. This function, $\sigma(t)$, is defined as:

```{=tex}
\[ \sigma(t) = \frac{1}{1 + e^{-t}} \]
```

The logistic function produces an S-shaped sigmoid curve, which asymptotically approaches 1 as $t$ increases and 0 as $t$ decreases. In logistic regression, the input $t$ is modeled as a linear combination of predictors (including an intercept), expressed as $\beta_0 + \beta_1X_1 + \beta_2X_2 + \dots + \beta_nX_n$. This transforms the predictors into probabilities of the binary outcome.

For this analysis, the predictors are employment status, income level, and living environment. These variables are passed through the logistic function to estimate $P(Y_i=1)$, the probability that respondent $i$ votes for Joe Biden.

Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013].

## Model Specification

The logistic regression model is specified as follows:

```{=tex}
\[
\log\left(\frac{P(Y_i=1)}{1 - P(Y_i=1)}\right) = \beta_0 + \beta_1X_{\text{employment},i} + \beta_2X_{\text{income},i} + \beta_3X_{\text{living},i}
\]
```

## Model Setup
$Y_i$: Binary indicator of support for Joe Biden (1) or Donald Trump (0) for respondent $i$.
$X_{\text{employment},i}$: Employment status of respondent $i$.
$X_{\text{income},i}$: Income level of respondent $i$.
$X_{\text{living},i}$: Living environment of respondent $i$.
$\beta_0$: Intercept of the model.
$\beta_1$, $\beta_2$, $\beta_3$: Coefficients quantifying the impact of employment, income, and living environment, respectively, on the likelihood of supporting Joe Biden.
Model Fitting
The model was implemented using the stan_glm() function from the rstanarm package in R [@citeR, @Rstanarm]. This function applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors. The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.


### Model Justification

#### Predictors

Employment, income, and living environment are socio-economic factors that have significant influence on voting preferences, as discussed in the data section.

These variables are categorical, making logistic regression a suitable choice for analyzing the relationship with a binary outcome.

#### Complexity

The model avoids unnecessary complexity by focusing on key variables while maintaining interpretability.

#### Bayesian Framework

Priors ($\beta_0, \beta_k \sim \text{Normal}(0, 2.5)$) are weakly informative, ensuring coefficients are regularized without unduly constraining the model [@Gelman2008].

#### Visualizations

Visualizing the relationships between predictors and voting preferences justifies their inclusion. This is shown in @fig-proportions. The heatmap @fig-heatmap-employment-living shows the correlation between the factors.

\newpage

# References