---
title: "Younger age groups, women, and lower levels of education are more likely to vote Democractic"
subtitle: "A logistic regression analysis of demographics in the 2020 US Presidential Election"
author: 
  - Victor Ma
thanks: "Code and data are available at: https://github.com/bestmustard/Determinants-of-Political-Support-in-the-United-States."
date: 03/16/2024
date-format: long
abstract: "In this study, I investigate demographic patterns which influence political support for the US 2020 election, specifically focusing on whether demographics of people tend to vote for Joe Biden or Donald Trump. I use a logistic regression model with the variables age, gender, and education to predict which categories in each demographic have the highest chance of voting for Joe Biden. I find that the model accurately predicts the trend of younger people, women, and those with the lowest level of education being more likely to support Joe Biden."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
    
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dataverse)
library(knitr)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(dplyr)
library(patchwork)
library(ggplot2)
library(broom.mixed)
ces_data <- read_csv("../data/analysis_data/analysis_data.csv")
ces_data$voted_for <- as_factor(ces_data$voted_for)
ces_data$education <- factor(ces_data$education, levels = c("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"))
```

# Introduction

The 2020 United States presidential election was a landmark event that encapsulated a confluence of crises and societal shifts. Beyond the COVID-19 pandemic, which reshaped healthcare priorities and exposed systemic vulnerabilities, the election underscored debates about economic inequality, particularly as millions faced unemployment during the economic downturn [@Blustein2020; @BLS2020]. Climate change emerged as a defining issue, with voters increasingly prioritizing environmental policies in response to intensifying natural disasters [@IPCC2020].

Demographic shifts also played a critical role. The increasing influence of younger, more diverse voters highlighted generational and racial divides in political priorities [@Frey2020]. Meanwhile, changes in suburban voting patterns, particularly among college-educated women, illustrated the evolving dynamics of key swing areas [@Cook2020].

Another major factor was the unprecedented role of mail-in voting, which became a focal point for controversy amidst the pandemic and claims of election fraud. This shift in voting methods not only increased accessibility for millions but also fueled political and legal battles over election integrity [@NCSL2020]. Finally, the election was a litmus test for the resilience of democratic institutions in the face of misinformation, political polarization, and challenges to the legitimacy of results [@BenGurion2020; @Sunstein2020].

Analyzing how living environments, employment status, and income levels correlate with political alignment in 2020 offers deep insights into societal trends. Research shows that urban residents tend to favor liberal candidates, reflecting the diversity and progressive social policies often prioritized in cities [@bishop2019bigsort]. Conversely, rural voters strongly supported Donald Trump, consistent with longstanding trends emphasizing traditional values and conservative policies [@cramer2016politics].

Employment status also played a key role. Data from the Cooperative Congressional Election Study (CCES, 2020) revealed that students and retirees, often focused on issues like education and healthcare, leaned more toward Joe Biden, while full-time workers, prioritizing economic stability, showed mixed preferences based on industry and region [@cces2020].

Income disparities further highlighted political divides. Lower-income groups, grappling with economic inequality, largely supported Biden due to his policies on healthcare and social safety nets, as noted in Pew Research’s 2020 voter analysis. Higher-income voters, on the other hand, leaned Republican, particularly in rural and suburban settings, reflecting preferences for tax policies that align with their financial interests [@pew2020voteranalysis].

These findings display the imapct of personal circumstances and political preferences, mapping not only the political landscape of 2020 but also providing a deeper understanding of the evolving priorities shaping voter behavior.

In this study, I use a logistic regression model to predict which candidate—Joe Biden or Donald Trump—was more likely to receive votes based on factors such as living environment, employment status, and income levels. Logistic regression is particularly suited for this analysis as it models binary outcomes, allowing us to estimate the likelihood of a voter aligning with one candidate over the other. By focusing on these explanatory variables, the study provides valuable insights into how socioeconomic and regional demographics influenced voter preferences during one of the most polarized elections in modern history.

The primary dataset for this study is the 2020 Cooperative Congressional Election Study (CCES), a stratified sample survey administered by YouGov and sourced from the Harvard Dataverse [@cces2020]. 

Complementary insights are drawn from Pew Research Center’s 2020 voter analysis, which contextualizes demographic trends across key groups, and the Census Bureau’s Current Population Survey Voting and Registration Supplement, which provides aggregate statistics on voter turnout and demographic characteristics [@pew2020; @CPS2020]. These additional datasets validate findings from the CCES and enhance the robustness of our conclusions.

My report is structured into five main sections and includes an appendix. After the introduction, the first section describes the CCES 2020 dataset, highlighting key attributes and presenting graphs of demographic distributions. The second section details the logistic regression model, its rationale, and preliminary findings. The third section analyzes how living environment, employment status, and income influenced election outcomes using visualizations and numerical results. The fourth section discusses the implications of the findings, potential weaknesses, and future research directions.

The appendix provides an in-depth exploration of survey methodologies, focusing on sampling and observational data related to the CCES 2020 dataset. It includes detailed discussions of idealized methodologies, literature comparisons, and simulations to illustrate sampling variability and bias, expanding on points in the Data section.

This analysis was conducted in R, utilizing several packages to streamline data handling, visualization, and modeling. Key tools include tidyverse for data manipulation and visualization, dataverse for accessing the CCES dataset, knitr for generating the report, modelsummary for interpreting models, and rstanarm for Bayesian regression analysis [@citeR; @Dataverse; @Knitr; @Modelsummary; @Rstanarm; @Tidybayes; @Dplyr]. Additionally, ggplot visualizations and elements of the "Discussion" section were developed with the assistance of ChatGPT4 @chatgpt.
\newpage


# Data
The dataset for this analysis is the 2020 Cooperative Election Study (CCES) [@ces2020], accessed via the Harvard Dataverse. The 2020 iteration surveyed 61,000 respondents, capturing a broad spectrum of U.S. political opinions. The release includes detailed documentation and questionnaires, ensuring transparency and replicability. Vote validation was conducted by Catalist, a large-scale organization maintaining data on over 240 million U.S. individuals [@Catalist2017].

## Strengths

### Sample Size

Sources for @fig-respondents below: [@ANES2020],[@Pew2020],[@Gallup2020], [@ces2020].

```{r out.height="30%"}
#| label: fig-respondents
#| fig-cap: "Number of respondents for various 2020 election survey sources"
#| echo: false
#| warning: false
#| message: false

# Create a data frame with the number of respondents for each source
survey_data <- data.frame(
  Source = c("ANES 2020", "Pew Research", "Edison Exit Polls", "CCES 2020"),
  Respondents = c(8280, 11818, 25000, 61000) # Updated with additional sources
)

# Generate the bar graph
ggplot(survey_data, aes(x = Source, y = Respondents, fill = Source)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Number of Respondents per Survey Source",
    x = "Survey Source",
    y = "Respondents"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Angle the x-axis text for readability

```


@fig-respondents shows that CCES poll features the largest respondent pool compared to other surveys. This extensive sample size enables granular analyses, ensuring the representation of smaller subgroups that may be underrepresented in smaller datasets [@Ansolabehere2020].

### Data Collection Phases
The CCES collected data in two waves, immediately before and after the 2020 election. This helps capture the shifts in voter attitudes and preferences during the critical final stages of the election campaign [@ces2020]. Pew Research and Gallup each conduct multiple polls of much smaller scale leading up to the election, which will give a snapshot in time of the voter sentiment but does not capture as broad of a perspective as the CES survey [@Pew2020], [@Gallup2020].

### Sampling Methodology
The CCES employs a two-step sample matching methodology designed to create a representative survey sample:

#### Target Sample Creation
A target sample is generated to reflect the broader population based on key demographic variables such as age, gender, race, and educational attainment. This target sample serves as a benchmark for representativeness.

#### Matched Sample Selection
For each individual in the target sample, a closely matched respondent is selected from a pool of survey volunteers maintained by YouGov. Matches are determined using a combination of publicly available databases and YouGov’s proprietary data, ensuring close alignment on variables such as geography, party registration, and voter history.

Once the matched sample is established, it is weighted to adjust for discrepancies between the sample and the target population. The weighting process involves two stages:

#### Demographic Weighting
Adjustments are made using data from the American Community Survey to ensure accurate representation across demographics.

#### Voting Behavior Weighting
Further adjustments incorporate validated voter registration data, accounting for turnout and voting patterns at the state and national levels.

Validation of the sampling methodology is conducted by comparing state-level survey results with official election outcomes. This comparison ensures that survey estimates align with actual voter behavior, providing confidence in the dataset's representativeness.

The sampling approach is specifically designed for online opt-in panels and mitigates biases commonly associated with such surveys, such as overrepresentation of politically engaged respondents. Unlike traditional random or quota sampling, this methodology emphasizes precision in matching and post-stratification to minimize systematic error [@ces2020], [@sample2007].

## Limitations
The data has some limitations. An error affected 925 North Carolina respondents who were shown incorrect House race candidates, impacting analyses of these districts. Additionally, self-reported data can introduce biases such as recall or social desirability bias. Pre-election surveys may also fail to capture last-minute shifts in voter sentiment.

The online nature of the survey, hosted by YouGov, expands accessibility but may underrepresent groups with limited internet access or technological literacy. Furthermore, the lack of an interviewer in online surveys could impact response quality [@ces2020].

## Variables of Interest
The selection of living environment, employment status, and income level as predictor variables for analyzing political preferences in the 2020 CCES data is guided by prior research that underscores their influence on voting behavior.

### Urban vs. Rural Residency
Urban and rural divides are well-documented predictors of political preferences. Urban voters have consistently leaned Democratic, driven by higher population density and exposure to diverse cultural and socioeconomic dynamics. For example, Pew Research found that 62% of urban voters supported Hillary Clinton in the 2016 election, compared to only 35% in rural areas [@Pew2018]. Rural voters, in contrast, exhibit stronger support for Republican candidates, influenced by traditional values and economic concerns rooted in agriculture and resource-based industries [@Cramer2016].

Research also shows that suburban areas, often politically contested, have shifted in recent years. The Cook Political Report highlights a notable swing in suburban support toward Democratic candidates in 2020, attributed to changes in demographics and education levels among suburban voters [@Cook2020].

### Income Level
Income levels are closely tied to political preferences, with higher-income individuals generally favoring Republican candidates due to tax policies, while lower-income groups often lean Democratic, prioritizing social welfare and redistribution policies [@Frank2004]. For example, the Census Bureau's Current Population Survey shows that in 2020, households earning under $50,000 were more likely to support Joe Biden, while those earning over $100,000 favored Donald Trump [@CPS2020].

However, this trend is nuanced by education levels and geographic factors. High-income earners in urban areas often prioritize social liberalism and climate change policies, aligning with Democratic platforms, while rural high-income earners focus more on fiscal conservatism [@Pew2020; @Edsall2020].

### Employment Level
Employment status significantly impacts voter preferences, with distinct patterns emerging across different occupational categories. Full-time workers are often divided along industry lines, with white-collar employees tending to support Democratic candidates and blue-collar workers aligning more with Republican candidates [@Muro2020]. Unemployment during economic crises, such as the COVID-19 pandemic, has further shaped voting patterns. Studies show that unemployed individuals are more likely to support candidates promising expansive social safety nets and job creation [@Blustein2020; @Fowler2020].

Retirees also exhibit unique voting behaviors, often prioritizing stability and healthcare, leading to a higher likelihood of Republican support [@KFF2020]. In contrast, students, who are less economically established and more progressive, tend to favor Democratic candidates [@Frey2020].

## Data Preparation and Cleaning
The 2020 CCES data, as recorded by Schaffner, Brian et al., was obtained from the Harvard Dataverse and processed using the arrow package for efficient storage in parquet format [@arrow]. The raw data was imported from a CSV file, and the cleaning process involved filtering for registered voters who cast votes for either Joe Biden or Donald Trump, treating the presidential vote as a binary outcome.

Key variables were transformed for clarity and analysis:

Living Environment: Categorized into "City," "Suburb," "Town," and "Rural area" based on the urbancity variable.
Employment Status: Labeled as "Full-time," "Part-time," "Temporarily laid off," "Unemployed," "Retired," "Permanently disabled," "Homemaker," or "Student."
Income Levels: Grouped into ranges: "< 10k," "10-50k," "50-100k," "100-200k," "200-500k," and "> 500k."
These transformations were implemented using the dplyr package for data manipulation. Each variable was converted into a factor with ordered levels to facilitate analysis and visualization.

The cleaned dataset was saved in both CSV and parquet formats for ease of use in subsequent analyses. Distributions for key explanatory variables, such as living environment, employment status, and income, are visualized in @fig-living, @fig-employment, and @fig-income below.

```{r out.height="20%"}
#| label: fig-living
#| fig-cap: "Number of respondents by area of living"
#| echo: false
#| warning: false
#| message: false

# Plot for area of living, excluding NA values
ggplot(cleaned_data %>% filter(!is.na(living)), aes(x = living)) +
  geom_bar(fill = "#87CEEB") +
  theme_minimal() +
  labs(title = "Respondents per Area of Living",
       x = "Living Environment",
       y = "Count") +
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 14))

```
```{r out.height="20%"}
#| label: fig-employment
#| fig-cap: "Number of respondents by employment status"
#| echo: false
#| warning: false
#| message: false

# Plot for employment status
ggplot(cleaned_data %>% filter(!is.na(employment)), aes(x = employment)) +
  coord_flip() +
  geom_bar(fill = "#87CEFA") +
  theme_minimal() +
  labs(title = "Respondents per Employment Status",
       x = "Employment Status",
       y = "Count") +
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 14))

```
```{r out.height="20%"}
#| label: fig-income
#| fig-cap: "Number of respondents by family income level"
#| echo: false
#| warning: false
#| message: false

# Plot for income level, excluding NA values
ggplot(cleaned_data %>% filter(!is.na(income)), aes(x = income)) +
  geom_bar(fill = "#FFD700") +
  theme_minimal() +
  labs(title = "Respondents per Income Level",
       x = "Income Level",
       y = "Count") +
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x-axis labels for clarity
```
@fig-living shows that the most common respondents were living in the suburbs and the city. 

In @fig-employment, the most common employment levels were full-time employees or retired.

@fig-income shows that the most common income levels were between the 10-100k range, and there was a high number of participants with an income level above $500,000.

```{r out.height="20%"}
#| label: fig-employmentvotes
#| fig-cap: "Proportion of votes by employment status"
#| echo: false
#| warning: false
#| message: false

# Plot for proportion of votes by employment status, excluding NA values
ggplot(cleaned_data %>% filter(!is.na(employment)), aes(x = employment, fill = voted_for)) +
  coord_flip() +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Proportion of Votes by Employment Status", 
       x = "Employment Status", 
       y = "Proportion")
```

@fig-employmentvotes shows clear trends in voting preferences. Students and those with lower income levels leaned more towards Joe Biden, likely reflecting generational and policy concerns such as healthcare and education funding. Conversely, individuals in full-time or part-time employment were more evenly split, with blue-collar occupations often skewing towards Donald Trump due to economic policies favoring traditional industries.
```{r out.height="30%"}
#| label: fig-scatter-income-employment
#| fig-cap: "Income vs. Employment Status"
#| echo: false
#| warning: false
#| message: false

ggplot(cleaned_data, aes(x = employment, y = as.numeric(income), color = voted_for)) +
  geom_jitter(width = 0.2, height = 0) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Income vs. Employment Status",
       x = "Employment Status",
       y = "Income Level",
       color = "Voted For")

```

@fig-scatter-income-employment shows us the income distribution across employment types.

Full-time workers exhibit the widest income distribution, spanning from "<10k" to ">500k."
Retirees and students predominantly occupy lower income levels, aligning with expectations based on fixed or limited income sources.

Retirees and students show a strong preference for Biden, while higher-income full-time workers lean slightly toward Trump.
Homemakers and permanently disabled individuals display narrower income ranges but include both Biden and Trump voters.

In the ">500k" category, Trump garners more support, consistent with Republican tax policies favoring high-income earners.
These patterns underscore the nuanced relationships between employment, income, and political preferences, highlighting the diversity of economic profiles within voter bases.

```{r out.height="30%"}
#| label: fig-heatmap-employment-living
#| fig-cap: "Heatmap of employment status and living environment"
#| echo: false
#| warning: false
#| message: false

# Pre-compute counts for the heatmap
heatmap_data <- cleaned_data %>%
  count(living, employment)

# Create the heatmap
ggplot(heatmap_data, aes(x = living, y = employment, fill = n)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() +
  labs(title = "Employment Status by Living Environment",
       x = "Living Environment",
       y = "Employment Status",
       fill = "Count")
```

The @fig-heatmap-employment-living provides insights into the distribution of employment statuses across different living environments:

Suburbs have the highest concentration of full-time workers, as indicated by the deep blue color. This aligns with the demographic trends of suburban areas, which often attract working professionals due to proximity to urban centers and family-friendly amenities.

Retirees are present in all living environments, with notable concentrations in both suburbs and rural areas. This reflects retirement trends where individuals seek quieter or more affordable living conditions outside cities.

Rural areas show a higher proportion of homemakers compared to other environments, consistent with traditional family roles in less urbanized areas.

Cities have the largest concentration of students, likely due to the presence of educational institutions and urban amenities.

Rural areas show higher proportions of unemployed and permanently disabled individuals compared to cities and suburbs, highlighting economic disparities and limited job opportunities in these regions.

Part-time work and temporary layoffs are distributed fairly evenly across living environments, suggesting they are less influenced by geographic location.

# Model

Logistic regression is a statistical model designed for binary outcomes, making it ideal for this study as we model votes for Joe Biden (1) versus Donald Trump (0).

The model calculates the log odds of a respondent voting for Joe Biden and maps it to a probability between 0 and 1 using the logistic function. This function, $\sigma(t)$, is defined as:

```{=tex}
\[ \sigma(t) = \frac{1}{1 + e^{-t}} \]
```

The logistic function produces an S-shaped sigmoid curve, which asymptotically approaches 1 as $t$ increases and 0 as $t$ decreases. In logistic regression, the input $t$ is modeled as a linear combination of predictors (including an intercept), expressed as $\beta_0 + \beta_1X_1 + \beta_2X_2 + \dots + \beta_nX_n$. This transforms the predictors into probabilities of the binary outcome.

For this analysis, the predictors are employment status, income level, and living environment. These variables are passed through the logistic function to estimate $P(Y_i=1)$, the probability that respondent $i$ votes for Joe Biden.

Logistic regression is well-suited for categorical and binary-dependent variables, and the explanatory variables in this study are categorical, making it an appropriate choice [@Hosmer2013].

## Model Specification

The logistic regression model is specified as follows:

```{=tex}
\[
\log\left(\frac{P(Y_i=1)}{1 - P(Y_i=1)}\right) = \beta_0 + \beta_1X_{\text{employment},i} + \beta_2X_{\text{income},i} + \beta_3X_{\text{living},i}
\]
```
## Model Setup
$Y_i$: Binary indicator of support for Joe Biden (1) or Donald Trump (0) for respondent $i$.
$X_{\text{employment},i}$: Employment status of respondent $i$.
$X_{\text{income},i}$: Income level of respondent $i$.
$X_{\text{living},i}$: Living environment of respondent $i$.
$\beta_0$: Intercept of the model.
$\beta_1$, $\beta_2$, $\beta_3$: Coefficients quantifying the impact of employment, income, and living environment, respectively, on the likelihood of supporting Joe Biden.
Model Fitting
The model was implemented using the stan_glm() function from the rstanarm package in R [@citeR, @Rstanarm]. This function applies Bayesian logistic regression with default priors and estimates the $\beta$ coefficients for the predictors. The model was fit to a subset of 3000 respondents from the 2020 CCES data processed earlier.


### Model Justification

In order to better interpret the results of the model, I can create a coefficient plot to visually see the effect sizes of the predictor variables on the likelihood of an individual supporting Biden. 

@fig-coefficient maps each predictor variable on the y-axis to an effect size and confidence interval on the x-axis. The effect size is the change in log-odds of supporting Biden for a one-unit increase in the predictor variable, which is essentially how much impact each demographic characterstic has an effect on voting for Biden. 

The confidence intervals tell us the range within we can be confident that the true effect lies. Smaller confidence intervals means there is a higher level of precision in the estimate of the effect size. I am not as interested in the intervals that cross zero because that means that there is data to support each side (Biden or Trump) and so they are less statistically significant. 


```{r}
#| label: fig-coefficient
#| fig-cap: "Coefficient plot of demographics"
#| echo: false
#| warning: false
#| message: false
political_preferences <- readRDS(file ="../models/political_preferences.rds")

# Assuming `tidy_political_preferences` is a tidy version of your model with terms and estimate
tidy_political_preferences <- tidy(political_preferences)

ggplot(tidy_political_preferences, aes(x = term, y = estimate)) +
  geom_point(aes(color = estimate > 0), size = 3) +  # color points based on positive or negative coefficient
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.2) +  #error bars
  coord_flip() +  
  scale_color_manual(values = c("blue", "red")) +  # Set colors for positive and negative coefficients
  labs(title = "Joe Biden Support",
       x = "Predictor Variables",
       y = "Effect on Log Odds of Supporting Biden",
       color = "Effect Direction") +
  theme_minimal()  # Use a minimal theme

# Ensure the 'tidy_political_preferences' contains the columns you want to summarize
tidy_political_preferences <- broom::tidy(political_preferences)

# Check available columns
names(tidy_political_preferences)
```

The conclusions I can draw from @fig-coefficient align with what I expected given the dataset. It's clear that higher education levels, specifically 4-year degrees and post-grad degrees, are positively associated with support for Biden. Higher age groups, specifically 45-64 and 65+, are strongly tied to voting for Trump. Men are less likely to vote for Biden than Women. 

\newpage
# Results

Figures @fig-page, @fig-pgender, and @fig-pedu are recreations of the earlier graphs we saw showcasing the proportion of voters for Joe Biden by groups within each demographic, where the bar is the data from CCES 2020 while the point is the prediction generated by the model. 

```{r out.height="25%"}
#| label: fig-page
#| fig-cap: "Model prediction for votes by age vs. CCES 2020 data"
#| echo: false
#| warning: false
#| message: false
# Example constant values for the other factors
constant_education <- "2-year" # Choose a representative value
constant_gender <- "Male" # Choose a representative value

# Create a data frame for age predictions
age_data <- data.frame(age = c("18-29", "30-44", "45-64", "65+"),
                       education = constant_education,
                       gender = constant_gender)

# Predict probabilities for each age group
age_data$predicted_prob <- predict(political_preferences, newdata = age_data, type = "response")
library(ggplot2)
# Actual proportions from the data
ces_data_grouped <- ces_data %>%
  group_by(age) %>%
  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))

# Assuming 'political_preferences' is your logistic regression model object and
# 'age_data' contains the predicted probabilities with 'age' as a factor
age_data$age <- factor(age_data$age, levels = c("18-29", "30-44", "45-64", "65+"))

# Create a combined plot
ggplot() +
  geom_bar(data = ces_data_grouped, aes(x = age, y = Proportion_for_Biden, fill = age), 
           stat = "identity", position = "dodge", alpha = 0.5) +
  geom_point(data = age_data, aes(x = age, y = predicted_prob, color = age), 
             position = position_dodge(width = 0.9), size = 3) +
  scale_fill_brewer(palette = "Set1") + 
  scale_color_brewer(palette = "Set1") + 
  labs(title = "Actual vs. Predicted Support for Biden by Age Group",
       x = "Age Group", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none") 

```

```{r out.height="25%"}
#| label: fig-pgender
#| fig-cap: "Model prediction for votes by gender vs. CCES2020 data"
#| echo: false
#| warning: false
#| message: false
# Example constant values for the other factors (choose representative values for each)
constant_age <- "30-44" # Assuming this is a middle category for age

# Create a data frame for gender predictions
gender_data <- data.frame(
  gender = c("Male", "Female"),
  education = constant_education, # Assuming you have a constant value set for education
  age = constant_age
)

# Predict probabilities for each gender category
gender_data$predicted_prob <- predict(political_preferences, newdata = gender_data, type = "response")

# Create a data frame for education predictions
education_data <- data.frame(
  education = c("No HS", "High school graduate", "Some college", "2-year", "4-year", "Post-grad"),
  gender = constant_gender, # Assuming you have a constant value set for gender
  age = constant_age
)

# Predict probabilities for each education category
education_data$predicted_prob <- predict(political_preferences, newdata = education_data, type = "response")

# Actual proportions by gender
ces_data_gender_grouped <- ces_data %>%
  group_by(gender) %>%
  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))

# Actual proportions by education
ces_data_education_grouped <- ces_data %>%
  group_by(education) %>%
  summarize(Proportion_for_Biden = mean(voted_for == "Biden", na.rm = TRUE))
# Plot for Gender
ggplot() +
  geom_bar(data = ces_data_gender_grouped, aes(x = gender, y = Proportion_for_Biden, fill = gender), 
           stat = "identity", position = "dodge", alpha = 0.5) +
  geom_point(data = gender_data, aes(x = gender, y = predicted_prob, color = gender), 
             position = position_dodge(width = 0.9), size = 3) +
  scale_fill_brewer(palette = "Set2") + 
  scale_color_brewer(palette = "Set2") + 
  labs(title = "Actual vs. Predicted Support for Biden by Gender",
       x = "Gender", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r out.height="25%"}
#| label: fig-pedu
#| fig-cap: "Model prediction for votes by education level vs. CCES2020 data"
#| echo: false
#| warning: false
#| message: false
# Plot for Education
ggplot() +
  geom_bar(data = ces_data_education_grouped, aes(x = education, y = Proportion_for_Biden, fill = education), 
           stat = "identity", position = "dodge", alpha = 0.5) +
  geom_point(data = education_data, aes(x = education, y = predicted_prob, color = education), 
             position = position_dodge(width = 0.9), size = 3) +
  scale_fill_brewer(palette = "Set3") + 
  scale_color_brewer(palette = "Set3") + 
  labs(title = "Actual vs. Predicted Support for Biden by Education",
       x = "Education Level", y = "Proportion") +
  theme_minimal() +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Adjust text angle for readability

```


By observation, it is evident that the model's prediction was fairly close to what the CCES data showed. The trend remains the same for each result. However, gender and education were predicted to have a higher sentiment of voting for Joe Biden while age had a lower sentiment. 

It is worth nothing though that in @fig-page, all age categories showed polling data of above 50% to vote for Biden. Now with the lower average vote, the older age categories of "45-64" and "65+" are leaning towards Trump. On @fig-pedu for the lower education groups, the "No-HS" group which was at just about 50% is now above it, and the "High School Graduate" is also closer to voting for Biden. While this may seem marginal, close elections where voter sentiment is close to 50% would be largely impacted by changes like this. Although ultimately this is not an accurate predictor for who would win the election, only which groups would vote for who.

# Discussion

This analysis demonstrates the demographics play in political preferences, with results that align with the existing literature. Our data suggests that age, education, and gender significantly influence electoral outcomes. In particular, the model's predictions closely mirrored CCES 2020 data trends, reinforcing the importance of these variables in understanding voter behavior. High education and younger ages showed high correlation with voting for joe biden.

It is important to note that based on the sample size of our data (as seen in @fig-education and @fig-agegender), the majority age groups which participated in the survey were "45-64" and "65+", while the variables of education and gender were much more closer together. If this were representative of the actual population then it would suggest that the age groups are skewed more towards those that support Trump, so despite most of the proportions being towards Biden, it can be a lot more nuanced of there are more groups in demographics that support Trump. 

The demographic trends observed in our analysis, specifically the tendency of younger voters, women, and individuals with higher educational attainment towards supporting Joe Biden, mirror broader societal shifts. These findings align with historical patterns where demographic attributes have increasingly become predictors of political alignment. For instance, the growing political polarization along educational lines has been well-documented, with individuals possessing higher educational qualifications tending to support more progressive policies and candidates [@Dalton2020].

Moreover, the gender gap in political preferences is not a novel phenomenon. Women have been consistently more likely to vote for Democratic candidates in recent elections, a trend that has been attributed to differing policy priorities, including healthcare, reproductive rights, and social welfare [@Schaffner2018]. The 2020 election further highlighted this divide, as issues such as women's rights and gender equality took center stage, prompting a significant mobilization among female voters [@Coffe2020].

Age demographics have also played a critical role in shaping the political landscape. The engagement and political leanings of younger voters have been subjects of extensive study, particularly given their inclination towards progressive and liberal policies. This has been partly driven by generational differences in value systems, with younger cohorts exhibiting greater concern for issues such as climate change, social justice, and economic inequality [@Norris2021]. The 2020 election underscored the importance of the youth vote, which showed overwhelming support for Biden, reflecting a broader trend of political mobilization among younger Americans in response to current socio-political challenges [@Smith2020].

The implications of these demographic trends extend beyond electoral outcomes. They signal shifting societal values and priorities, with potential long-term impacts on policy direction and political discourse. For instance, the increasing support for progressive policies among younger and more educated demographics may influence the policy agendas of political parties and candidates in future elections [@Inglehart2020].

It's also crucial to note the limitations of relying solely on demographic variables to predict electoral outcomes. While age, education, and gender provide significant insights, the complex interplay of socio-economic factors, geographic location, and individual experiences also plays a crucial role in shaping political preferences. Future research should aim to incorporate a broader range of variables to capture the complexity of voter behavior more accurately.

Our analysis contributes to the growing body of literature on demographic influences on political preferences, offering valuable insights into the dynamics of the 2020 U.S. presidential election. As the nation continues to navigate through periods of significant social and political change, understanding the demographic underpinnings of electoral outcomes will remain a critical area of study.

# Weaknesses and Next Steps
Although valuable, this study provides an intriguing analysis of how demographic influences affect political preferences; however, there are additional limitations to consider. Firstly, the study's analysis model, which used logistic regression, may have simplified the understanding of complex dynamics and could have overlooked some interactions between demographic variables that would provide a better understanding of voter behavior. Future research could utilize more advanced tools, such as multilevel or mixed-effects models, to more accurately capture the interplay of these demographic variables.

Secondly, while detailed, the study's dataset may not fully account for the subtleties in voter behavior across all demographics. For example, the impact of regional urban-rural political contexts on voting patterns could have nuances beyond the current analysis's reach. Incorporating geographic and socioeconomic variables could further improve and strengthen the model, offering deeper insight into the electoral impact of demographic factors.

Lastly, the study's focus on the 2020 election, though timely, limits its ability to capture longitudinal trends in voter behavior. A comparative analysis developed over multiple election cycles might reveal changes in demographic influences on political preferences over time, providing a richer context for interpreting the 2020 election results. In summary, this paper not only sheds light on the demographic determinants of voter preferences in a pivotal election year but also encourages reflection on the broader societal shifts these patterns indicate. As America faces deep social and political challenges, understanding the demographic underpinnings of political allegiance becomes increasingly important. Further research, building on the insights provided in this paper and addressing the weaknesses outlined, has the potential to offer even better explanations of the complex tapestry of American electoral politics.
\newpage

# References